---
title: Importance of red cards
author: R package build
date: '2021-04-04'
slug: Importance-of-red-cards
categories:
  - R
tags:
  - EPL
  - red cards
  - soccer
---
Besides goals, there are few things in soccer that change games quite like a red card. Late in an important match, the last thing that any manager wants is one of their players being sent off, reducing their team to ten men, or at least, that's what common sense would say. However red cards do not automatically equate a loss and quite often teams manage to hang on to tie or even win while down a player. So let's take a look at just how costly red cards are, and are earlier red cards more detrimental than later red cards?  

```{r}
library(tidyverse)
```

I collected my data from eplreview.com and used the 2017-18, 18-19, and 19-20 English Premier League red card statistics to guide me to the answer. It included the player, the team, the game, the minute, and the final result of every red card which was very helpful.

```{r}
realfootball<- read_csv("EPLredcard_data.csv")
```

From there, I narrowed it down to the time the red card was issued, and what the result of the game was for that team. and added whether it was a win (W), a loss (L), or a tie (T). 

```{r}
realfootball %>% 
  group_by(`Time of Red Cards Given`, WLT) %>%
  summarise(
    total = n())%>% 
  arrange(`Time of Red Cards Given`)
  
```

After this I created a data set with just the data I needed and nothing else. This allowed me to focus on adding more data from the data I had and clean it to make it more understandable.

```{r}
RCD <- realfootball %>%
  group_by(`Time of Red Cards Given`, WLT)%>%
  summarise(
    total = n())
RCD
```


This was the tricky bit. After sorting out the time, result, and total, I separated the regular time and additional time after the 45th minute so that red cards during added time would be represented in their respective halves. What this means is red cards that occurred in the 45+5 minute still counts as a red card in the 45th minute and not the 50th minute due to the nature of the game. Now we can see "RegularTime" and "AddedTime" which helps account for the red cards during added time while still counting it as part of the half it occurred in. 

From there, I sorted every red card into 6 groups based on when it it occured in the match. All red cards from the first 15 minutes became one data set, and every 15 minute increment for the rest of the match. I felt that the 6 data points would be more digestible for the readers rather than 90 different data points. I could go with five minute increments as well but I chose 15 because I felt it would be easier to comprehend and there was more data per increment for 15 rather than five. I didn't do ten minute increments because I feel that red cards from the 45th minute or before halftime should not be comparable to red cards from after halftime, so 15 minute increments was perfect.

```{r}
RCD %>% 
  separate(`Time of Red Cards Given`, into=c("RegularTime", "AddedTime"), sep="\\+") %>% 
  mutate_all(funs(ifelse(is.na(.), 0, .))) %>% 
  mutate(
    `RegularTime` = as.numeric(gsub("'", "", `RegularTime`, fixed=TRUE)),
    `AddedTime`   = as.numeric(gsub("'", "", `AddedTime`, fixed=TRUE)),
    total_time = `RegularTime` + `AddedTime`
    ) %>% 
  mutate(
    game_section = case_when(
      RegularTime <= 15  ~ "1-15",
      RegularTime >= 16 & RegularTime <= 30 ~ "16-30",
      RegularTime >= 31 & RegularTime <= 45~ "31-45",
      RegularTime >= 46 & RegularTime <= 60~ "46-60",
      RegularTime >= 61 & RegularTime <= 75 ~ "61-75",
      RegularTime >= 76 & RegularTime <= 90 ~ "76-90",
      TRUE ~ as.character(total_time)
      ) 
  ) -> footballredcards

```

For the graph I stacked all the red cards from their corresponding section into one column and sorted it by whether the team won, lost or tied. this shows us the results of every team who recieved a red card over the season  and how they fared in the match based on when the red card occurred.

As we can see, In every section from the first minute to the 60th minute, the majority of teams who recieved a red card went on to lose the game. However, from the 61st minute to the 75th minute, teams actually had a better chance of winning and tying than they did of losing. In the 76th minute to the 90th minute we see a return to what we would expect with more losses than wins or ties again. 


```{r}
ggplot(footballredcards, aes(x=game_section, weight=total, fill=WLT)) + 
  geom_bar() +
  labs(
    title="Results based on Timing of Red Cards in the Premier League", 
    x="Minute of Red Card Given", 
    y="Amount of Red Cards Given",  
    subtitle="Just how detrimental are red cards actually?.",
    caption="Source: eplreview.com | By Joseph Reina") +
  theme_minimal()+
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.title = element_text(size = 12), 
    plot.subtitle = element_text(size=12), 
    panel.grid.minor = element_blank()
    )
```

This next chart shows a little better just how many more red cards were given in the final fifteen minutes compared to the rest of the game. Just under 30% of all red cards were issued during the final fifteen minutes, nearly double what we would expect if every red card was issued evenly throughout the game. 

```{r}
ggplot() + geom_bar(data=footballredcards, aes(x=game_section, weight=total, fill=WLT)) + theme_minimal() + coord_polar()+
  labs(
    title="Results based on Timing of Red Cards in the Premier League", 
    x="Minute of Red Card Given", 
    y="Amount of Red Cards Given",  
    subtitle="Just how detrimental are red cards actually?.",
    caption="Source: eplreview.com | By Joseph Reina") +
  theme_minimal()+
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.title = element_text(size = 12), 
    plot.subtitle = element_text(size=12), 
    panel.grid.minor = element_blank()
    )
```

With this in mind, a reason that so many teams in the final 30 minutes went on to win or tie might be due to a player making a rash tackle or decision to preserve a lead and the opposing team did not have enough time to effectively make use of their numeric advantage.

In the end, playing with 10 players definitely hurts a team's chances of winning, and red cards in the first 60 minutes are the biggest culprit.
